---
title: "TV Product Conjoint Analysis"
author: "Shruthi Khurana"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Executive Summary

This analysis uses conjoint analysis to determine optimal product positioning and pricing for a TV product. Key findings:

- **Optimal Price**: $2,400
- **Maximum Profit**: $45,583
- **Market Share**: 70% (below $2,500), 42% (at/above $2,500)
- **Recommended Product**: Sony 4K 75-inch TV

Resolution quality is the most important feature (40.65% importance), followed by brand (23%) and screen size (23.78%). Price sensitivity is relatively low (12.4%), indicating willingness to pay for premium features.

# Load Function

```{r load-function}
source("conjoint_analysis.R")
```

# Run Analysis

```{r run-analysis, fig.width=12, fig.height=6}
# Run conjoint analysis
results <- conjoint_analysis(
  preferences_file = "data/Design_Matrix.xlsx",
  
  # Own product: Sony 4K 75" at low price
  own_design = c(
    "Resolution 4K = 1" = 1,
    "Sony = 1" = 1,
    "Price (low = 0; hi =1)" = 0,
    "Screen 75 inch" = 1,
    "Screen 85 inch" = 0
  ),
  
  # Competitor A: Sony 4K 75" at high price
  compA_design = c(
    "Resolution 4K = 1" = 1,
    "Sony = 1" = 1,
    "Price (low = 0; hi =1)" = 1,
    "Screen 75 inch" = 1,
    "Screen 85 inch" = 0
  ),
  
  # Competitor B: Non-Sony 4K 75" at low price
  compB_design = c(
    "Resolution 4K = 1" = 1,
    "Sony = 1" = 0,
    "Price (low = 0; hi =1)" = 0,
    "Screen 75 inch" = 1,
    "Screen 85 inch" = 0
  ),
  
  # Feature costs: [4K, Sony, HighPrice, 75", 85"]
  feature_costs = c(1000, 500, 1000, 250, 250)
)
```

# Results

## Part-Worth Utilities

Part-worth utilities show how much value each feature adds to overall product preference. The baseline (intercept) represents a 65-inch Sharp non-4K TV at low price.

```{r partworth}
knitr::kable(
  results$Partworth_Table,
  digits = 2,
  caption = "Part-Worth Utility Estimates"
)
```

**Key Insights:**

- **4K Resolution (+3.83)**: Strongest positive impact on preference
- **Sony Brand (+2.17)**: Moderate brand premium
- **75" Screen (~0)**: Minimal preference vs. 65" baseline
- **85" Screen (-2.25)**: Negative preference for oversized screen
- **High Price (-1.17)**: Price sensitivity penalty

## Attribute Importance

Attribute importance measures the relative influence of each feature category on overall preference.

```{r importance}
imp_pct <- results$Attribute_Importance * 100
knitr::kable(
  data.frame(
    Attribute = c("Screen Size", "Resolution", "Brand", "Price"),
    Importance = c(imp_pct$Screen, imp_pct$Resolution, imp_pct$Brand, imp_pct$Price)
  ),
  digits = 2,
  caption = "Attribute Importance (%)"
)

# Visualization
barplot(
  as.numeric(imp_pct),
  names.arg = c("Screen", "Resolution", "Brand", "Price"),
  main = "Attribute Importance",
  ylab = "Importance (%)",
  col = c("#3498db", "#e74c3c", "#2ecc71", "#f39c12"),
  las = 1
)
```

**Preference Hierarchy**: Resolution (40.65%) > Brand (23%) ≈ Screen Size (23.78%) > Price (12.4%)

This indicates a quality-focused consumer who prioritizes display technology and brand trust over cost savings.

## Willingness to Pay

Willingness to pay (WTP) converts utility values into dollar amounts, showing how much consumers would pay for each feature upgrade.

```{r wtp}
knitr::kable(
  data.frame(
    Feature = c("75\" Screen", "85\" Screen", "4K Resolution", "Sony Brand"),
    WTP = c(
      results$Willingness_to_Pay$Screen_75,
      results$Willingness_to_Pay$Screen_85,
      results$Willingness_to_Pay$Resolution_4k,
      results$Willingness_to_Pay$Brand_Sony
    )
  ),
  digits = 0,
  caption = "Willingness to Pay ($)"
)
```

**Key Findings:**

- **4K Resolution**: $1,637 premium (highest value driver)
- **Sony Brand**: $928 premium (strong brand loyalty)
- **75" Screen**: ~$0 (indifferent to mid-size upgrade)
- **85" Screen**: Negative WTP (would require discount)

## Market Share

Market share predictions using multinomial logit model:

```{r market-share}
knitr::kable(
  results$Market_Share,
  digits = 2,
  caption = "Predicted Market Share"
)
```

**At optimal price ($2,400):**
- Own product captures **70% market share** (below $2,500 threshold)
- Competitor A: ~15% (same features, higher price)
- Competitor B: ~15% (different brand, same price)

## Pricing Optimization

```{r optimal-price}
cat(sprintf("Optimal Price: $%d\n", results$Optimal_Price))
cat(sprintf("Maximum Profit: $%,.2f\n", results$Maximum_Profit))
cat(sprintf("Market Share at Optimal Price: 70%%\n"))
```

**Profit Curve Insights:**

- **Below $2,000**: Losses (below cost)
- **$2,000**: Break-even
- **$2,400**: Maximum profit at $45,583
- **≥$2,500**: Steep profit decline (price penalty triggers, market share drops to 42%)

The $2,500 threshold represents a psychological pricing barrier that causes dramatic market share loss.

# Strategic Recommendations

## Product Positioning

**"Premium Quality at Accessible Pricing"**

Offer a **Sony 4K 75-inch TV priced at $2,400** that:

1. **Delivers high-importance features** (Resolution: 40.65%, Brand: 23%)
2. **Stays below psychological threshold** ($2,500 triggers steep share loss)
3. **Captures quality-focused segment** (low price sensitivity: 12.4%)
4. **Maximizes profit** while maintaining 70% market share

## Target Customer

**"The Quality Seeker"**

- Values display technology (4K resolution) above all else
- Trusts established brands (Sony reputation)
- Willing to pay for premium features
- Less concerned with absolute price levels
- Prefers optimal viewing experience over maximum screen size

## Competitive Advantages

| Factor | Advantage |
|--------|-----------|
| **High Resolution** | Delivers 40.65% of decision weight |
| **Sony Brand** | Commands $928 premium in customer minds |
| **Below $2,500** | Avoids -28pp market share drop |
| **70% Share** | Dominates competitive set |
| **$45,583 Profit** | Maximizes financial performance |

## Implementation

1. **Product Development**: Prioritize 4K technology over screen size expansion
2. **Pricing**: Set at $2,400, never cross $2,500 threshold
3. **Marketing**: Lead with resolution quality and Sony brand trust
4. **Portfolio**: Offer 65" and 75" models, skip 85" (negative preference)

# Appendix: Visualizations

The plots generated during analysis show:

1. **Market Share vs. Price**: Sharp drop at $2,500 threshold
2. **Profit vs. Price**: Clear peak at $2,400 optimal price

These visualizations validate the pricing strategy and highlight the importance of staying below the $2,500 psychological barrier.

---

*Analysis completed: `r Sys.Date()`*  
*Course: BAx 442 - Advanced Statistics, UC Davis*
